---
title: "Introduction to Hidden Markov Models"
author: "Jared Stabach, Smithsonian's National Zoo & Conservation Biology Institute"
date: '`r format(Sys.time(), "%d %B %Y")`'
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: true
      smooth_scroll: true
    number_sections: false
    #theme: united
    #highlight: tango
pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<a href="https://github.com/Smithsonian/Wildebeest_HMM.git" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#151513; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>

# Introduction
  >Many thanks to Dr. Joe Kolowski for assistance in developing this material
  
In this exercise, we will fit Hidden Markov Models (HMMs) with the [moveHMM](https://cran.r-project.org/web/packages/moveHMM/index.html) package to assess changes in wildebeest behavior.  We will continue to use the dataset from previous exercises that has been cleaned and subset to the Athi-Kaputiei Plains study area in southern Kenya.  Remember that the data have been re-sampled to a 3-hour fix interval.  We'll investigate the role of temperature and time of day on the movement behaviors of wildebeest, but many other environmental variables (e.g., Anthropogenic Disturbance) could be added to further our understanding on potential impacts to animals.

All code is based on exercises developed by Joe Kolowski at the Smithsonian-Mason School of Conservation, with sections modified from the [moveHMM](https://cran.r-project.org/web/packages/moveHMM/index.html) guide. 

In this exercise, we will:

  * Learn the analysis workflow for fitting Hidden Markov Models with [moveHMM](https://cran.r-project.org/web/packages/moveHMM/index.html)
  * Practice investigation of impact of continuous covariates on state transition probabilities 
  * Learn how to interpret output from HMMs
  * Practice methods for visualizing HMM output both spatially and graphically
  * Use model output to subset tracking datasets
  
## Load Libraries
Load the required libraries and remove everything held in [R's](https://cran.r-project.org/) memory.

```{r Setup, message=FALSE, warning=FALSE, echo=TRUE}
# Remove items from memory/clean your workspace
rm(list=ls())

# You may need to install these packages first
#install.packages('moveHMM', 'ggmap', 'tidyverse', )

# Load libraries
library(moveHMM)
#library(ggmap)
library(tidyverse)
```

## Data Import & Formatting
Load the cleaned wildebeest dataset from the previous exercise.   In terms of data preparation, a few processing/formatting steps are required:

  * It is assumed that location error is relatively small. If you are working with a high error data set, you should likely process it before using this package. For example, you may try a state space approach, such as [crawl](https://cran.r-project.org/web/packages/crawl/index.html) or [aniMotum](https://ianjonsen.github.io/aniMotum/articles/Overview.html). 
  * Note that [moveHMM](https://cran.r-project.org/web/packages/moveHMM/index.html) does not use date/time information.  We have already verified that our dataset is regular, but you also need to verify that the data is organized chronologically.
  * Variables included in analyses must be numeric.  For example, seasons should be changed to 1, 2, 3, and 4 to be incorporated in analyses. 
  * The animal id column must be named "ID".
  * It seems very helpful to convert the coordinates (currently in meters) to km by dividing by 1000.  All [moveHMM](https://cran.r-project.org/web/packages/moveHMM/index.html) vignettes do this unit conversion.

```{r Load, message=FALSE, warning=FALSE, echo=TRUE}
# Read the dataset into R, selecting the necessary columns (x, y, date, id) for analyses.
# We'll also grabe the temperature column, the sex of each animal, and create an hour field from the date field
# Lastly, we will convert the distance move data to km (e.g., x/1000)
WB.data <- read_rds("Data/wildebeest_3hr_adehabitat.rds") %>%
  arrange(id, date) %>%
  select(x,
         y,
         t = date,
         ID = id,
         temp,
         sex) %>% 
  mutate(hour = lubridate::hour(t),
         x = x/1000,
         y = y/1000,
         sex = as.numeric(sex))
```

# Create Move Object
Now that we have imported the data, we will use the function `prepData` to prepare the data for analyses.  If our data were raw lat/long values, we could still import the data, but we would need to set the type = "LL" (default). See `help(prepData)` for additional information.  

**Note** This package only works with longitude/latitude and UTM projected data.

```{r Create, message=FALSE, warning=FALSE, echo=TRUE}
# Create Object
WB.move <- WB.data  %>% 
  prepData(type = "UTM",
           coordNames = c("x","y"))

# What's the class of the object?
class(WB.move)
```

### Object Summary
The [moveHMM](https://cran.r-project.org/web/packages/moveHMM/index.html) package includes a number of functions to help summarize the content of the move object.  We'll see that we have movement data for all 12 wildebeest (a good check!), with two covariates (temperature and hour of day) to use in analyses.  We also see from above that the format is a data.frame, making it relatively easy to modify our object if we need to.  Also notice that the function `prepData` created two additional columns in our dataset, "step (steplength)" and "angle (turning angle)".  These are the two fields that will be used in all future modeling steps with this package.  The plot function will show the movement trajectory and include plots of the steplengths and turning angles. 

```{r Move Summary, message=FALSE, warning=FALSE, echo=TRUE}
# Summarize the object
summary(WB.move)

# What are the column headings included in this move object?
names(WB.move)

# We can plot all animals together.  Please try!
# plot(WB.move,
#      compact = TRUE, # Make sure you specify compact = TRUE.  If you don't specify, each plot will be drawn separately, which can be tedious
#      ask = FALSE)

# Or, each animal separately
# unique(WB.move$ID) # We can always query the dataframe for the names of each animal.
plot(WB.move[WB.move$ID == "Kikaya",],
     ask = FALSE)

# We can also investigate the steplength and turning separately, using standard R commands
# hist(WB.move$step)
# summary(WB.move$step)
# quantile(WB.move$step,
#          probs = 0.90,
#          na.rm = TRUE)
# hist(WB.move$angle)
```

# Fitting a Hidden Markov Model
Start at line 106

## Set Starting Values